Liveness-Based Garbage Collection for Lazy Languages

### 2016-02-08 paper/paper.tex

\definecolor{Myblue}{rgb}{.2,0,1}
\definecolor{Myred}{rgb}{1,.2,.2}
\newcommand{\mycomment}[1]{{\color{Myblue}{#1}} }
%\newcommand{\added}[1]{{\color{Myblue}{#1}}}
\newcommand{\added}[1]{#1}
\newcommand{\todelete}[1]{}%{\color{Myred}{#1}}}

@@

We next  describe the function $\mathcal{L}$  that propagates liveness
across  expressions.   Consider  the $\mathcal{L}$-rules  for  {\LET},
{\SIF}, and {\SRETURN}.  Since the value of $(\SRETURN~x)$ is the
value  of  $x$,  a  demand $\sigma$  on  $(\SRETURN~x)$  gives  a
liveness   of  $\{x.\sigma\}$.    The  liveness   of  the   expression
$(\SIF~x~e_1~e_2)$  is  a union  of  the  liveness of  $e_1$  and
$e_2$. In  addition, since  the condition $x$  is also  evaluated, the
liveness $\{x.\epsilon\}$ is created and  added to the union.  
To  understand  the  liveness rule  for  $\LET~x  \leftarrow~s~\IN~e$,
observe that the value  of $\LET$ is the value of  its body $e$.  Thus
the  liveness environment  $\Lv$ of  $e$ is  calculated for  the given
demand $\sigma$. Since the stack variable $x$ is copied to each of the
closure variable  $x_{\pi}$, the liveness of  $x$ is the union  of the
liveness of the  closure variables.  This liveness,  say $\sigma'$, is
also  the  demand  on  $s$, thus the  liveness  environment
$\mathit{ref}(s,   \sigma',   \Lfonly)$   is  added   to   $\Lv   \cup
\{x.\sigma\}$. Finally, the stack variables corresponding to the free
variables of $s$ are updated and added 
  to  give the  overall liveness environment  for $(\LET~x
\leftarrow~s~\IN~e)$.

As noted earlier,  $x.\alpha \in \Lv_y$ specifies the  liveness of the
reference $\heap\llbracket\rho(x), \alpha\rrbracket$  only if $\alpha$
corresponds to a closure-free path  in \heap\ starting from $\rho(x)$.
If this path is intercepted by a closure, say $(\CAR~{x_{\pi}})$, then
the liveness of heap contents starting from $\mathit{x_{\pi}}$
is given by $\Lv_\mathit{x_{\pi}}$.

 As we  shall see in section??,  the liveness of the  closure variable
 $x_{\pi}$ is recorded along with the closure so that the GC can refer
 to it during garbage collection.

%} %comment

@@

\item  Consider the following environments, 1) environment in the
   current state, 2) environments in each of the stacked items and 3)
   the environment in each of the closures in the heap. For each of
   these environments we can calculate a corresponding liveness environment
   using the  $e\ and\ \sigma$. 
\item \emph{GC}  for each  $x \in domain(\rho)$,  we set  $\rho(x)$ to
  $\bot$ iff $x.\epsilon \notin \Lv$.  Similarly, for each stack entry
  $(\rho,\ell,e,\sigma)$  in  $S$  with $\Lv'$  as  the  corresponding
  liveness environment in  $\mathsf{SL}$, and finally for  each $x \in
  domain(\rho)$  for some  closure in  the heap,  we set  $\rho(x)$ to
  $\bot$ iff $x.\epsilon \notin \Lv'$ corresponding to the closure.

@@

\begin{lemma}
\label{lemma:call-less-cannot-go-bang}
Consider the minefield execution of  a program without function calls.
Such a program cannot enter the \bang\ state.
\end{lemma}
\begin{proof}
Call a $GC(\ldots)$  followed by a $\rightsquigarrow$  transition as a
step. Consider a state $(\rho, S,  \heap, e, \sigma)$ in the minefield
execution of a  program.  We show by induction on  the number of steps
$n$ leading  to this  state that  the next  transition cannot  enter a
\bang\ state.  When $n$ is 0,  the state is $([\;]_\rho,\,    (\rho_\mathit{init},   \, \ell_{\ans}  ,\,
(\print~\ans)):[\;]_{S}, 
[\;]_{\heap},\, (\mainpgm))$.   Since  the
            call to $GC$  in this state does  nothing, we just
            have to show that the $\rightsquigarrow$ transition cannot
            enter  a \bang\  state.  Since  our programs  are in  ANF,
            $e_{\mainpgm}$ can  only be  a $\LET$ expression.   A {\sc
              let}  step  does  not involve  dereferencing,  and  thus
            cannot result in a \bang.

>>>>

\definecolor{Myblue}{rgb}{.2,0,1}
\definecolor{Myred}{rgb}{1,.2,.2}
\newcommand{\mycomment}[1]{{\color{Myblue}{#1}}}
\newcommand{\added}[1]{{\color{Myblue}{#1}}}
%\newcommand{\added}[1]{#1}
\newcommand{\todelete}[1]{{\color{Myred}{#1}}}

@@

We next  describe the function $\mathcal{L}$  that propagates liveness
across  expressions.   Consider  the $\mathcal{L}$-rules  for  {\LET},
{\SIF}, and {\SRETURN}.  Since the value of $(\SRETURN~x)$ is the
value  of  $x$,  a  demand $\sigma$  on  $(\SRETURN~x)$  gives  a
liveness   of  $\{x.\sigma\}$.    The  liveness   of  the   expression
$(\SIF~x~e_1~e_2)$  is  a union  of  the  liveness of  $e_1$  and
$e_2$. In  addition, since  the condition $x$  is also  evaluated, the
liveness $\{x.\epsilon\}$ is created and  added to the union.  
To  understand  the  liveness rule  for  $\LET~x  \leftarrow~s~\IN~e$,
observe that the value  of $\LET$ is the value of  its body $e$.  Thus
the  liveness environment  $\Lv$ of  $e$ is  calculated for  the given
demand $\sigma$. Since the stack variable $x$ is copied to each of the
closure variable  $x_{\pi}$, the liveness of  $x$ is the union  of the
liveness of the  closure variables.  This liveness,  say $\sigma'$, is
also  the  demand  on  $s$, thus the  liveness  environment
$\mathit{ref}(s,   \sigma',   \Lfonly)$   is  added   to   $\Lv   \cup
\{x.\sigma\}$. Finally, the stack variables corresponding to the free
variables of $s$ are updated and added 
  to  give the  overall liveness environment  for $(\LET~x
\leftarrow~s~\IN~e)$.
{\color{Myblue}
As noted earlier,  $x.\alpha \in \Lv_y$ specifies the  liveness of the
reference $\heap\llbracket\rho(x), \alpha\rrbracket$  only if $\alpha$
corresponds to a closure-free path  in \heap\ starting from $\rho(x)$.
If this path is intercepted by a closure, say $(\CAR~{x_{\pi}})$, then
the liveness of heap contents starting from $\mathit{x_{\pi}}$
is given by $\Lv_\mathit{x_{\pi}}$.

 As we  shall see in section??,  the liveness of the  closure variable
 $x_{\pi}$ is recorded along with the closure so that the GC can refer
 to it during garbage collection.
}
%} %comment

@@

 \item  Consider  the following  environments,  1)  environment in  the
  current state, 2)  environments in each of the stacked  items and 3)
  the environment  in each of  the closures in  the heap. For  each of
  these  environments  we  can   calculate  a  corresponding  liveness
  environment using the $e\ and\ \sigma$.

  \item {\color{Myblue}\emph{GC} for  each $x \in domain(\rho)$, we  set $\rho(x)$ to
    $\bot$  iff $x.\epsilon  \notin \Lv$.   Similarly, for  each stack
    entry   $(\rho,\ell,e,\sigma)$  in   $S$   with   $\Lv'$  as   the
    corresponding liveness  environment in $\mathsf{SL}$,  and finally
    for  each  $x  \in  domain(\rho)$ for  some  closure  $\langle  s,
    \lfloor\rho\rfloor_{FV(s)}, \sigma\rangle$ in the heap, for all $x
    \in  FV(s)$ we  set $x$  to  $\bot$ iff  $x.\epsilon \notin  \Lv'$.}

@@

\begin{lemma}
\label{lemma:call-less-cannot-go-bang}
Consider the minefield execution of  a program without function calls.
Such a program cannot enter the \bang\ state.
\end{lemma}
\begin{proof}
Call a $GC(\ldots)$  followed by a $\rightsquigarrow$  transition as a
step. Consider a state $(\rho, S,  \heap, e, \sigma)$ in the minefield
execution of a  program.  We show by induction on  the number of steps
$n$ leading  to this  state that  the next  transition cannot  enter a
\bang\  state.    When  $n$   is  0,   the  state   is  $([\;]_\rho,\,
(\rho_\mathit{init},   \,  \ell_{\ans}   ,\,  (\print~\ans)):[\;]_{S},
        [\;]_{\heap},\, (\mainpgm))$.  Since the  call to $GC$ in this
        state  does   nothing,  we  ju1st   have  to  show   that  the
        $\rightsquigarrow$  transition cannot  enter  a \bang\  state.
        Since our  programs are in  ANF, $e_{\mainpgm}$ can only  be a
        $\LET$  expression.   A  {\sc   let}  step  does  not  involve
        dereferencing, and thus cannot result in a \bang.
